Investigating shape standardisation using geometric morphometry of pottery from Iron Age northeastern Taiwan

### 2020-04-10 analysis/paper/paper.Rmd
editing discussion: remove metrics

To compare pottery shape standardization across the three phases we investigate the distributions of the first three PC scores, taking the PC scores as proxy variables for vessel shape (Figure \@ref(fig:PCs-distribution)). The first PC, vessel height, shows higher variation in the pre-European period compared to the Chinese period. That indicates that shape standardization was higher in the Chinese period compared to the pre-European period. The second PC also presents a similar pattern of higher shape standardization in the Chinese period compared to the other two phases. To see whether the differences in the distribution of PCs between any two phases are substantive or due to chance, we used a modified signed-likelihood ratio test to assess the equality of CVs. P-values for this significance test of CVs for PC1 and PC2 show significant differences in the standardization of vessel shapes across periods, especially between Chinese contact with either pre-European or post-European (Figure \@ref(fig:CV-test-PCs-plot), Table \@ref(tab:CV-test-PCs-table)). A significant difference is also detected between the pre-European and post-European periods.

@@

Previous investigations at Kiwulan have suggested an unequal distribution of prestige goods, trade ornaments specifically, following the appearance of Europeans [@Cheng2008; @Wang2011], hinting at the emergence of social inequality in the Indigenous community. Here we have examined ceramic vessel shape standardization to measure craft specialization as a proxy for social differentiation, such as elite control [@Costin2001; @Junker1999]. Our results indicate that differences in pottery shape can be detected using EFA. The result of our MANOVA test shows there is a significant difference in shapes between the pre-European and Post-European periods, and between the pre-European and the Chinese periods. The average shape presents a round body with a wider rim and neck before European contact, and turns to a more oval-shaped body with narrower rim and neck after the European presence. These changes even more pronounced during the Chinese contact period. In general, pottery vessels become shorter in height over time, leading to an oval-shaped body.

Our CV tests indicate that there are significant differences in shape standardization between the pre-European period and post-European period, and between the Chinese contact and either pre-European or post-European periods. This suggests pottery shape became more homogeneous and standardized after foreign contacts with European colonizers and even more so after contact with Chinese immigrants. Compositional analysis shows that the clay pastes are similar, regardless of the increasing standardization of the pottery shape, reflecting continuity in the raw material sources. We can thus rule out changes in clay fabric as a factor in explaining changes in vessel shape. To determine if shape changes might be related to changes in the function of pots at Kiwulan, we used geochemical methods to extract and identify lipids trapped in the fabric of potsherds to identify foods that may have contributed residues absorbed into the clay [cf. @Kwak2015]. Unfortunately, we did not obtain useful results due to extremely low lipid yields, which is probably due to the very thin, dense, and low porosity fabric of Kiwulan pottery. These physical characteristics of the clay offer limited spaces to trap and protect organic molecules from microbiological degradation [cf. @Evershed2008, pp. 909].

We compare the result of shape variable with the result of metric data for six measurements and two ratio values using CV test (Table \@ref(tab:metricCVstest)). Metric measurements show there is a significant difference (p-value less than 0.01) between CV values for two metric measurements across phases, the rim diameter and the neck diameter before and after the European presence. After the arrival of the Chinese, there is only one metric attribute, the rim diameter, showing significant difference compared with post-European period (p-value less than 0.05). Although these results show the differences between different contact phases that corresponds to part of the results of shape variable, elliptic Fourier analysis captures more subtle differences in a substantial way reflected by overall shapes of pottery, not only limited to diameter variables. In addition, the result of spatial analysis (Figure \@ref(fig:kernel-plot)) for pottery shape samples presents multiple clusters with high density of pottery during the European presence. However, the hypothesis testing on spatial randomness indicates a non-randomly dispersed distribution before European contact and more extreme dispersed after European presence. In contrast, the distribution of pottery is more similar to the random distributions during the Chinese period. This contradicts our expectation that clustered pattern will be observed with an increase in pottery standardization since the emergence of specialized groups. Or, this may indicate that there are other factors lead to pottery shape standardization instead of a small number of artisans.

```{r metricCVstest}

# select metric attributes and tidy data for computing CV
kwl_p_metrics_long <-
  kwl_p_ave %>%
  mutate(dia_rim_body = Dia_Rim_ave/Dia_Body_ave,
         thi_rim_body = Thick_Rim_ave/Thick_Body_ave) %>%
  select(period, 
         Dia_Rim_ave, Dia_Neck_ave, 
         Dia_Body_ave, Thick_Rim_ave, 
         Thick_Neck_ave, Thick_Body_ave,
         dia_rim_body, thi_rim_body,) %>%
  rename(`Rim diameter (mm)` = Dia_Rim_ave,
         `Neck diameter (mm)` = Dia_Neck_ave,
         `Body diameter (mm)` = Dia_Body_ave,
         `Rim thickness (mm)` = Thick_Rim_ave,
         `Neck thickness (mm)` = Thick_Neck_ave,
         `Body thickness (mm)` = Thick_Body_ave,
         `Ratio of Rim/Body diameter` = dia_rim_body,
         `Ratio of Rim/Body thickness` = thi_rim_body) %>%
  gather(variable, value, -period) %>%
  filter(!is.na(period), !is.na(value)) %>% 
  mutate(variable = factor(variable, levels = c("Rim thickness (mm)",
                                                "Neck thickness (mm)",
                                                "Body thickness (mm)",
                                                "Ratio of Rim/Body thickness",
                                                "Rim diameter (mm)",
                                                "Neck diameter (mm)",
                                                "Body diameter (mm)",
                                                "Ratio of Rim/Body diameter"))) %>%
  arrange(variable)

# filter pre-e and post-e periods
kwl_p_metrics_long_pre_post <-
  kwl_p_metrics_long %>%
  filter(!period == "ch-con")
# filter post-e and chi periods
kwl_p_metrics_long_post_chi <-
  kwl_p_metrics_long %>%
  filter(!period == "pre-e")

#---------------------------CV test for pre-e and post-e pair
kiwulan_pottery_metrics_cvs_plot_cv_test_asym_pre_post <-
  kwl_p_metrics_long_pre_post %>%
  nest(-variable) %>%  # or nest(period, value) %>%
  mutate(asymptotic_test = map(data, # change from mslr_test to asymptotic_test
                               ~bind_cols(asymptotic_test (
                                 .x$value,
                                 .x$period)))) %>%
  unnest(asymptotic_test)
# compute CV
kwl_p_cvs_pre_post <-
  kwl_p_metrics_long_pre_post %>%
  group_by(period, variable) %>%
  summarise(cvs = raster::cv(value, na.rm = TRUE))
# join p-values to CVs
kwl_p_cvs_and_pvalues_pre_post <-
  kwl_p_cvs_pre_post %>%
  left_join(kiwulan_pottery_metrics_cvs_plot_cv_test_asym_pre_post) %>%
  mutate(significant = ifelse(p_value <= 0.05, 'yes', 'no')) %>%
  mutate(variable = factor(variable, ordered = TRUE)) %>%
  arrange(variable)
# making table for CV
kwl_p_cvs_table_pre_post <-
  kwl_p_cvs_and_pvalues_pre_post %>%
  select(-data, -significant) %>%
  pivot_wider(names_from = period, values_from = cvs) %>%
  select(variable, `pre-e`, `post-e`, `D_AD`, `p_value`) %>%
  mutate_at(c("pre-e", "post-e", "D_AD", "p_value"), round, 4)

#---------------------------CV test for post-e and ch-con pair
kiwulan_pottery_metrics_cvs_plot_cv_test_asym_post_chi <-
  kwl_p_metrics_long_post_chi %>%
  nest(-variable) %>%  # or nest(period, value) %>%
  mutate(asymptotic_test = map(data, # change from mslr_test to asymptotic_test
                               ~bind_cols(asymptotic_test (
                                 .x$value,
                                 .x$period)))) %>%
  unnest(asymptotic_test)
# compute CV
kwl_p_cvs_pre_post <-
  kwl_p_metrics_long_post_chi %>%
  group_by(period, variable) %>%
  summarise(cvs = raster::cv(value, na.rm = TRUE))
# join p-values to CVs
kwl_p_cvs_and_pvalues_post_chi <-
  kwl_p_cvs_pre_post %>%
  left_join(kiwulan_pottery_metrics_cvs_plot_cv_test_asym_post_chi) %>%
  mutate(significant = ifelse(p_value <= 0.05, 'yes', 'no')) %>%
  mutate(variable = factor(variable, ordered = TRUE)) %>%
  arrange(variable)
# making table for CV
kwl_p_cvs_table_post_chi <-
  kwl_p_cvs_and_pvalues_post_chi %>%
  select(-data, -significant) %>%
  pivot_wider(names_from = period, values_from = cvs) %>%
  select(variable, `post-e`, `ch-con`, `D_AD`, `p_value`) %>%
  mutate_at(c("post-e", "ch-con", "D_AD", "p_value"), round, 4)

# join two table together
kwl_p_cvs_table_all <-
  kwl_p_cvs_table_pre_post %>% 
  left_join(kwl_p_cvs_table_post_chi, "variable") %>% 
  select(-`post-e.y`) %>% 
  rename(`before European` = "pre-e", `European` = "post-e.x", Chinese = "ch-con",
         `D_AD-1` = "D_AD.x", `D_AD-2` = "D_AD.y",
         `P value-1` = `p_value.x`, `P value-2` = `p_value.y`) %>% 
  select(variable, `before European`, European, Chinese,
         `D_AD-1`, `P value-1`, `D_AD-2`, `P value-2`)

knitr::kable(kwl_p_cvs_table_all,
             caption = "Coefficient of Variation for metric attributes by three phases using Asymptotic test [@Feltz1996]. The first set of P-value represents the comparison between before European and European periods, while the second represents the comparison between European and Chinese periods ")
```

>>>>

To compare pottery shape standardization across the three phases we investigate the distributions of the first three PC scores, taking the PC scores as proxy variables for vessel shape (Figure \@ref(fig:PCs-distribution)). The first PC, capturing vessel height and roundness, shows higher variation in the pre-European period compared to the Chinese period. That indicates that shape standardization was higher in the Chinese period compared to the pre-European period. The second PC also presents a similar pattern of higher shape standardization in the Chinese period compared to the other two phases.

To see whether the differences in the distribution of PCs between any two phases are substantive or due to chance, we used a modified signed-likelihood ratio test to assess the equality of CVs [citation needed: article and pkg]. P-values for this significance test of CVs for PC1 and PC2 show significant differences in the standardization of vessel shapes across periods, especially between Chinese contact with either pre-European or post-European (Figure \@ref(fig:CV-test-PCs-plot), Table \@ref(tab:CV-test-PCs-table)). A significant difference was also detected between the pre-European and post-European periods.

@@

Previous investigations at Kiwulan have suggested an unequal distribution of prestige goods, trade ornaments specifically, following the appearance of Europeans (Cheng, 2008; Wang, 2011), hinting at an emergence of social inequality in the Indigenous community. We have examined ceramic vessel shape standardization to measure craft specialization as a proxy for social differentiation (Costin, 2001; Junker, 1999). The result of our MANOVA test shows there is a significant difference in shapes between the pre-European and Post-European periods, and between the pre-European and the Chinese periods. The average shape presents a round body with a wider rim and neck before European contact, and turns to a more oval-shaped body with narrower rim and neck after the European presence. These changes even more pronounced during the Chinese contact period. In general, pottery vessels become shorter in height over time, leading to an oval-shaped body.

Our CV tests indicate that there are significant differences in shape standardization between the pre-European period and post-European period, and between the Chinese contact and either pre-European or post-European periods. Pottery shape became more homogeneous and standardized after contact with Europeans and even more so after contact with the Chinese. Compositional analysis shows that the clay pastes are similar, regardless of the increasing standardization of the pottery shape, reflecting continuity in the raw material sources. We can thus rule out changes in clay fabric as a factor in explaining changes in vessel shape. To determine if shape changes might be related to changes in the function of pots at Kiwulan, we used geochemical methods to extract and identify lipids trapped in the fabric of potsherds to identify foods that may have contributed residues absorbed into the clay (cf. Kwak and Marwick, 2015). Unfortunately, we did not obtain useful results due to extremely low lipid yields, which is probably due to the very thin, dense, and low porosity fabric of Kiwulan pottery. These physical characteristics of the clay offer limited spaces to trap and protect organic molecules from microbiological degradation (cf. Evershed, 2008, p. 909).

The spatial analysis (Figure \@ref(fig:kernel-plot)) for pottery samples presents multiple clusters with high densities of pottery during European presence. Hypothesis testing on spatial randomness indicates a non-randomly dispersed distribution before European contact and more extreme dispersed distribution after European presence. In contrast, the distribution of pottery is more similar to random distributions during the Chinese period. This contradicts our expectation that clustered pattern will be observed with an increase in pottery standardization since the emergence of specialized groups. The absence of clusters in the Chinese period is notable because this was a time of a historically-documented decline of Indigenous population (Chen, 2007; Hsieh, 2009).  We might expect reduced numbers of potters to result in pottery production shrinking to a few locations in the settlement during this time. Although the overall number of vessels is smaller during the Chinese contact period, Figure 7) shows that pottery is distributed randomly across the sampling area without any distinctive clusters during this time. This suggests that differences in population across our three occupation phases are probably not driving variation in craft specialization.
